---
title: "Markov Chains, Concretely"
subtitle: "The backbone of diffusion models — first of a three-part refresher series"
description: "A worked introduction to Markov chains — transition matrices, stationary distributions, and the Gaussian forward chain at the heart of DDPM — verified with R simulation and closed-form checks."
author: "Paul Milne"
date: "2026-09-06"
categories: [explainer, diffusion, deep-learning]
engine: knitr
bibliography: references.bib
format:
  html:
    toc: true
---

This is the first of three short refreshers behind two earlier posts — [Reproducing a Phase Transition in Diffusion Models](/posts/diffusion-phase-transition/) and [Fixed Points, Phase Transitions, and Diffusion Models](/posts/rg-diffusion-phase-transitions/) — both of which lean on ideas from Sclocchi, Favero & Wyart (2024) [@sclocchi2024phase] that are easy to gloss over if you haven't seen them recently. This one covers Markov chains; the next two cover mean-field analysis and the generative diffusion process itself.

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
theme_set(theme_minimal(base_size = 12))
```

## What makes a process "Markov"

A process is Markov if its next state depends only on its *current* state — not on the path that got it there. Formally, $P(x_{t+1} \mid x_t, x_{t-1}, \dots, x_0) = P(x_{t+1} \mid x_t)$. That's it; the entire history collapses into whatever's true right now.

A simple weather model makes this concrete: three states, and a transition matrix giving the probability of tomorrow's weather given today's.

```{r}
#| label: fig-weather-path
#| fig-cap: "A 200-step simulated path through the weather chain, colour-coded by state."
states <- c("Sunny", "Cloudy", "Rainy")
P <- matrix(
  c(0.7, 0.2, 0.1,
    0.3, 0.4, 0.3,
    0.2, 0.3, 0.5),
  nrow = 3, byrow = TRUE, dimnames = list(states, states)
)

set.seed(1)
n_steps <- 200
path <- integer(n_steps)
path[1] <- 1L
for (t in 2:n_steps) path[t] <- sample.int(3, 1, prob = P[path[t - 1], ])

data.frame(step = seq_len(n_steps), state = factor(states[path], levels = states)) |>
  ggplot(aes(step, state, colour = state)) +
  geom_point(size = 1.5) +
  labs(x = "step", y = NULL, colour = NULL, title = "Simulated weather path") +
  theme(legend.position = "top")
```

Knowing today is Sunny tells you everything relevant about tomorrow; knowing it was also Sunny for the last five days adds nothing. That's the whole idea, and it's what makes chains like this tractable to analyze — no exploding state space of "all possible histories" to track.

## Where does it settle? Two ways to find out

Run the chain long enough and the fraction of time spent in each state converges to a fixed *stationary distribution* $\pi$, satisfying $\pi P = \pi$ — a left eigenvector of $P$ with eigenvalue 1. Two independent ways to find it should agree:

```{r}
#| label: fig-stationary
#| fig-cap: "Stationary distribution via eigen-decomposition vs. a long simulated path's empirical frequencies."
eig <- eigen(t(P))
pi_stat <- Re(eig$vectors[, which.min(abs(eig$values - 1))])
pi_stat <- pi_stat / sum(pi_stat)
names(pi_stat) <- states

n_long <- 5000
path_long <- integer(n_long)
path_long[1] <- 1L
for (t in 2:n_long) path_long[t] <- sample.int(3, 1, prob = P[path_long[t - 1], ])
pi_empirical <- table(factor(states[path_long], levels = states)) / n_long

df <- data.frame(
  state = rep(states, 2),
  probability = c(pi_stat, as.numeric(pi_empirical)),
  method = rep(c("eigenvector", "5000-step simulation"), each = 3)
)

ggplot(df, aes(state, probability, fill = method)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "probability", fill = NULL)
```

Eigenvector: Sunny `r round(pi_stat["Sunny"], 3)`, Cloudy `r round(pi_stat["Cloudy"], 3)`, Rainy `r round(pi_stat["Rainy"], 3)`. A long simulated path lands close to the same numbers — the chain has "forgotten" where it started and settled into its long-run behaviour. This idea — a distribution that's a *fixed point* of an update rule — will come back in the next post, where a self-consistency equation plays exactly the same role.

## DDPM's actual Markov chain

Diffusion models run a Markov chain too, just with a continuous state and Gaussian transition kernels rather than a lookup table. The forward (noising) process is defined, exactly as in Ho, Jain & Abbeel (2020) [@ho2020denoising], by:

$$
q(x_t \mid x_{t-1}) = \mathcal{N}\!\left(\sqrt{1-\beta_t}\, x_{t-1},\ \beta_t I\right)
$$

Each step nudges $x_{t-1}$ slightly toward the origin and adds a little Gaussian noise, with $\beta_t$ controlling how much. This is precisely the recurrence used in `guided_diffusion/gaussian_diffusion.py`, the codebase reproduced in [the diffusion-phase-transition post](/posts/diffusion-phase-transition/).

```{r}
#| label: fig-gaussian-chain
#| fig-cap: "Eight sample trajectories of the scalar Gaussian chain, starting from x0=2, collapsing into the N(0,1) envelope (shaded bands at +-1, +-2 sd) under the real Ho et al. (2020) schedule."
T_steps <- 1000
beta <- seq(1e-4, 0.02, length.out = T_steps)
alpha <- 1 - beta
alpha_bar <- cumprod(alpha)

x0 <- 2.0
n_traj <- 8
set.seed(2)
traj <- matrix(NA_real_, T_steps + 1, n_traj)
traj[1, ] <- x0
for (t in 1:T_steps) {
  eps <- rnorm(n_traj)
  traj[t + 1, ] <- sqrt(alpha[t]) * traj[t, ] + sqrt(beta[t]) * eps
}

traj_df <- as.data.frame(traj) |>
  mutate(step = 0:T_steps) |>
  tidyr::pivot_longer(-step, names_to = "trajectory", values_to = "x")

band_df <- data.frame(step = 0:T_steps, sd = sqrt(1 - alpha_bar[c(1, seq_len(T_steps))]))

ggplot() +
  geom_ribbon(data = band_df, aes(step, ymin = -2 * sd, ymax = 2 * sd), fill = "grey85") +
  geom_ribbon(data = band_df, aes(step, ymin = -sd, ymax = sd), fill = "grey70") +
  geom_line(data = traj_df, aes(step, x, group = trajectory), colour = "steelblue", linewidth = 0.4) +
  labs(x = "t", y = expression(x[t]), title = "Forward chain trajectories vs. their theoretical envelope")
```

## The closed-form jump

Simulating 1000 sequential steps every time you need a noisy sample would be wasteful, and because each step's kernel is Gaussian, it's also unnecessary: the composition of $t$ Gaussian steps is itself Gaussian, in closed form:

$$
q(x_t \mid x_0) = \mathcal{N}\!\left(\sqrt{\bar\alpha_t}\, x_0,\ (1-\bar\alpha_t) I\right), \qquad \bar\alpha_t = \prod_{s \le t}(1-\beta_s)
$$

You can jump straight from $x_0$ to any $x_t$ in one draw. This is exactly `q_sample` in the reproduced code, and it's what makes the "forward-diffuse-partway-then-reverse" experiment in [the diffusion-phase-transition post](/posts/diffusion-phase-transition/) practical on real images. Worth actually checking it's true rather than taking it on faith:

```{r}
#| label: closed-form-check
check_t <- 30
n_mc <- 200000
set.seed(3)
x_stepwise <- rep(x0, n_mc)
for (t in 1:check_t) {
  eps <- rnorm(n_mc)
  x_stepwise <- sqrt(alpha[t]) * x_stepwise + sqrt(beta[t]) * eps
}
mean_cf <- sqrt(alpha_bar[check_t]) * x0
var_cf <- 1 - alpha_bar[check_t]
```

At $t=`r check_t`$, 200,000 step-by-step Monte Carlo draws give mean `r round(mean(x_stepwise), 4)` and variance `r round(var(x_stepwise), 5)`; the closed form predicts mean `r round(mean_cf, 4)` and variance `r round(var_cf, 5)` — matching to within Monte Carlo noise.

## The reverse chain

The reverse process $p_\theta(x_{t-1} \mid x_t)$ is modelled the same way — Markov, approximately Gaussian for small $\beta_t$ — but its mean isn't known in closed form for real data; that's what a neural network is trained to approximate. [The next explainer](/explainers/mean-field-annealed-disorder/) is a detour into a different piece of machinery this paper needs (mean-field analysis); [the third](/explainers/toy-diffusion-model/) comes back to build a toy reverse chain where, unusually, the right answer *is* known exactly.

## References

::: {#refs}
:::
